---
title: "Day 2 Answer Sheet"
output:
  html_document:
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

strange_object <- "ghost!"
```

## Booleans / Logicals

```{r}
# Test if 3 - 2 = 4.
3 - 2 == 4
```


```{r}
# Test if strange_object is a character object.
is.character(strange_object) == TRUE
```


```{r}
age_in_months <- 74
age_in_years <- 6.1667
age_group <- "middle_school"

# Add your code below to query these objects and determine their variable type
typeof(age_in_months)
typeof(age_in_years)
typeof(age_group)
```


```{r}
school <- c("middle", "elementary", "middle", "middle", "high", "high", "primary", "elementary")

# Enter your code below to make this list/variable a factor and use is.factor to determine if you have succeeded


```


```{r}
library(tidyverse)
rm(list = ls())
school_directory <- read_csv("12_basics_data.csv")

# Enter your code below

school_directory$school <- as_factor(school_directory$school)

school_directory %>%
  mutate('home room' = as_factor('home room'))

```

## Operators and NA ----------------

How would you make a series of numbers from 4 to 12, named "series"?
```{r}
# Enter your code below.
series <- 4:12
```

Now write some code to test if "series" is greater than 10. 
What type of object is the output?

```{r}
# Enter your code below.
series > 10
```

I have a vector of city names called `cities`. How many NA values does it have?

```{r}
cities
```

Can you write code to test if the values in `cities` equal NA?
Remember, the operator to test for equality is `==`, NOT `=`.
```{r}
# Enter your code below.
cities == NA # this won't work :)
```

If you tried to write `cities == NA`, why didn't that work?
A special function for testing for NA values is `is.NA()`.

Try `is.NA()` with `cities`.
```{r}
# Enter your code below.
is.na(cities)
```

Now try writing code to test if values in `cities` are *not* NA.

```{r}
!is.na(cities)
```

