---
title: "Data Visualization Lab"
output: html_notebook
---

Check out this cheat sheet for using ggplot2:
https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf

Now this lab will just be you exploring data using ggplot!

Be creative! Try different things. 

We have some datasets you can use:
- The CPI dataset we used today
- A dataset of all of the eruptions that have occurred worldwide in the Holocene (the current era lol)
- A dataset of ASL signs from ASL-LEX

Import the datasets using this chunk, or use the `got_data` df, or your own data.

```{r}
cpi_tidy <- readRDS("data/cpi_tidy.rds")
eruptions <- readRDS("data/holocene_eruptions.rds")
sign_data <- readRDS("data/asl_lex_sign_data.rds")
```

If you want to start with what we already made in class, here it is!

```{r}
# Compare the prices of different food items from 2020-2023
food_items <- c("Meats", "Eggs", "Fruits and vegetables", "Sugar and sweets")

cpi_foods <- cpi_tidy %>% 
  filter(item %in% food_items)

ggplot(cpi_foods,
       mapping = aes(
         x = year,
         y = increase,
         group = item
       )) +
  geom_line() +
  geom_point() + 
  labs(title = "% Increase in Food Prices 2020-2023",
       x = "Year",
       y = "% Increase") +
  theme_classic()
```


```{r}
# Summarize average annual increase in food prices from 2020-2023
library(tidyverse)
ggplot(cpi_tidy,
       mapping = aes(
         x = year,
         y = increase
         )) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point") +
  labs(title = "Average % Increase in Food Prices 2020-2023",
       x = "Year",
       y = "% Increase",
       color = "Food Type") +
  theme_classic()
```

Or you can try making a ggplot from scratch!

You can add a chunk by pressing `CMD-OPTION-I` or the green `+C` icon at the top. Have fun!

```{r}
# Make your own plots below!

```

