---
title: "Manipulating Data"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Let's import the dataset for `eruptions`:

```{r}
eruptions <- readRDS("data/holocene_eruptions.rds")
```

## Arranging and Filtering Rows ----------

Let's view the `eruptions` dataset.

```{r}
View(eruptions)
```


We can see it's arranged by volcano number. What if you wanted to arrange it by year?
Assign this to a new object called `eruptions_chronological`

```{r}
# Arrange eruptions by year in a 
# Write your code below


```


Now maybe you want to only see eruptions in a specific country.
Choose a country and select for that country only. Assign this to `eruptions_chronological_yourcountry`.

```{r}
# Look at chronological eruptions 
# Write your code below
```


Can you *start* with the `eruptions` dataset and try to use arrange() and filter() like above, but using the pipe so that you don't have to assign new objects every time?
I've started your code block below. Try to finish this code to get the same dataset as `eruptions_chronological_yourcountry`.
No need to assign this to a new object.

```{r}
# Arrange eruptions in one country by year

eruptions %>% # Write your code below
  

```

## Selecting and Mutating Columns ----------

The `eruptions` dataset has too many columns/variables! It's annoying to scroll through columns to see what variables you're interested in.
Let's make a dataset `eruptions_smaller` that includes only the following columns:
  - volcano_name
  - year
  - country
  - elevation

Feel free to add any other variables you're interested in!
  
```{r}
# Select columns of interest in `eruptions`.
# Write your code below


```


Great! However, most of us are Americans! We don't understand the concept of "meters". 
Let's create a new variable for elevation in feet, named `elevation_ft`.
Assign this to a new object `eruptions_smaller_americanized`.
FYI: the conversion rate for meters to feet is 1 meter = 3.28 feet.

```{r}
# Help Americans understand elevation data in `eruptions`.
# Write your code below



```



## Recoding Variables ----------

You can also mutate a new column where you *selectively* change values to other values, using `case_match()` in `mutate()`.
This is perfect for recoding variables!

For example, maybe you want to recode "female" to 0, "male" to 1, "nonbinary" to 2, and so on.

Or for `eruptions`, you want to recode `activity_evidence` to a new variable: `seen`, meaning did people see the eruption?

`activity_evidence` has three values: "Eruption Dated", "Eruption Credible", "Eruption Observed". 
However, the first two both mean people did not SEE the eruption, so in our new variable they would both be `no`.

Here is how you would do that using `case_match`. I am adding this to the *original* dataset, not assigning it a new object.
This means I am rewriting the original dataset so I better be sure it works!

```{r}
# Create a new variable for underwater vs. above ground eruptions
eruptions <- eruptions %>% 
  mutate(seen = case_match(
    activity_evidence,
    "Eruption Dated" ~ "no",
    "Eruption Credible" ~ "no",
    "Eruption Observed" ~ "yes"
  )
)
```


